---
title: "Spatial Transcriptomics – Image Registration and Multiomics Integration"
author: "Ding Yang Wang"
date: "2025-04-04"
output:
  html_document: default
  pdf_document: default
output_dir: ../../docs/module2
editor_options: 
  markdown: 
    wrap: 72
---

<nav style="background-color: #f8f9fa; padding: 10px; font-family: sans-serif; border-bottom: 1px solid #ddd;">

<a href="../index.html" style="margin-right: 20px; text-decoration: none;">
← Back to Note</a>
<a href="https://github.com/Dewey-Wang/Compgen-2025" target="_blank" style="text-decoration: none;">
View on GitHub</a>

</nav>

```{=html}
<style>
.center {
  display: flex;
  justify-content: center;
  align-items: center;
  text-align: center;
}
</style>
```

# Description

This R Markdown document demonstrates how to: - Perform **image
registration** to align transcriptomic and histological data. -
Integrate **cell-level and molecule-level spatial transcriptomics**
(Xenium). - Combine **tissue morphology** from H&E with **molecular
insights** to study spatial infection patterns in COVID-19 lungs.

We will use: - Xenium breast cancer dataset with accompanying H&E
image. - A Xenium COVID-19 lung dataset with annotated hyaline membrane
regions and SARS-CoV-2 viral RNAs (`S2_N`, `S2_orf1ab`).

> If you want to reproduce the experiment: 👉 [Click here to the Docker
> and the data setup instructions](#environment--dataset-setup)

## Description of the Files

-   `Xen_R1`: Processed Xenium object of breast cancer section (cells).
-   `Xen_R1_image`: Corresponding H&E image.
-   `vr2_merged_acute1`: COVID-19 Xenium dataset with both **cells** and
    **molecules** (`Xenium` and `Xenium_mol`).
-   `acutecase1_membrane.geojson`: Annotated hyaline membrane regions.
-   `acutecase1_heimage.jpg`: H&E image of lung tissue.

```{r setup, include=FALSE}
options(java.parameters = "-Xmx8g")
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

# Load Required Libraries

```{r load-libraries, message=FALSE}
library(VoltRon)
library(dplyr)
library(Seurat)
library(ComplexHeatmap)
```

------------------------------------------------------------------------

# Image Registration – Xenium Breast Cancer Dataset

## Why Image Registration?

Spatial omics technologies capture molecular and histological views
independently. **Image registration** allows us to align: -
**Transcriptomic data (Xenium DAPI)** and - **Morphological data (H&E
images)**

This enables **morphology-guided cell analysis**, e.g., identifying
tumor boundaries or immune cell infiltration zones.

## Import Data

\\{r} Xen_R1 \<- loadVoltRon("../data/ondisk/Xen_R1/") \\\

## Register with H&E

\\{r} Xen_R1_image \<-
importImageData("../data/BreastCancer/Xenium_R1/Xenium_FFPE_Human_Breast_Cancer_Rep1_he_image_highres.tif",
sample_name = "HEimage", channel_names = "H&E", tile.size = 100)
Xen_R1_image_disk \<- saveVoltRon(Xen_R1_image, format = "HDF5VoltRon",
output = "../data/ondisk/Xen_R1_image/", replace = TRUE)
Xen_R1_image_disk \<- loadVoltRon("../data/ondisk/Xen_R1_image/") \\\

### Perform Registration

\\{r} xen_reg \<- registerSpatialData(object_list = list(Xen_R1,
Xen_R1_image_disk)) Xenium_reg \<- xen_reg\$registered_spat[[2]]
vrImages(Xen_R1[["Assay1"]], channel = "H&E") \<- vrImages(Xenium_reg,
name = "main_reg", channel = "H&E") vrSpatialPlot(Xen_R1, group.by =
"Clusters", channel = "H&E") \\\

------------------------------------------------------------------------

# Multi-omics Integration – COVID-19 Lung

## Biological Background

COVID-19 lungs contain viral RNAs (`S2_N`, `S2_orf1ab`), some inside
cells, some in **hyaline membrane** regions (mucosal damage zones). We
analyze: - **Cell types**: Immune and structural cells. - **Molecules**:
Viral RNA transcripts.

## Import Xenium Object

\\{r} vr2_merged_acute1 \<- readRDS(file =
"../data/COVID19/acutecase1_annotated.rds")
SampleMetadata(vr2_merged_acute1) \\\

## Visualize Cells & Molecules

\\{r} \# Cells vrSpatialPlot(vr2_merged_acute1, assay = "Xenium",
group.by = "CellType", plot.segments = TRUE)

# Molecules

vrSpatialPlot(vr2_merged_acute1, assay = "Xenium_mol", group.by =
"gene", group.ids = c("S2_N", "S2_orf1ab"), n.tile = 500) \\\

### Insight

Molecules often lie **outside cells**, particularly in damaged zones,
making tissue morphology crucial for interpretation.

------------------------------------------------------------------------

# H&E Alignment with ROI

## Import & Register H&E Image

\\{r} imgdata \<-
importImageData("../data/COVID19/acutecase1_heimage.jpg", tile.size =
10, segments = "../data/COVID19/acutecase1_membrane.geojson",
sample_name = "acute case 1 (HE)", channel_names = "H&E") imgdata \<-
flipCoordinates(imgdata, assay = "ROIAnnotation") imgdata \<-
modulateImage(imgdata, brightness = 300) xen_reg \<-
registerSpatialData(object_list = list(vr2_merged_acute1, imgdata)) \\\

## Transfer Image and ROI

\\{r} imgdata_reg \<- xen_reg\$registered_spat[[2]]
vrImages(vr2_merged_acute1[["Assay7"]], name = "main", channel = "H&E")
\<- vrImages(imgdata_reg, assay = "Assay1", name = "main_reg")
vrImages(vr2_merged_acute1[["Assay8"]], name = "main", channel = "H&E")
\<- vrImages(imgdata_reg, assay = "Assay1", name = "main_reg")

# Add ROI Assay

vr2_merged_acute1 \<- addAssay(vr2_merged_acute1, assay =
imgdata_reg[["Assay2"]], metadata = Metadata(imgdata_reg, assay =
"ROIAnnotation"), assay_name = "ROIAnnotation", sample = "acute case 1",
layer = "Section1") vrMainSpatial(vr2_merged_acute1[["Assay9"]]) \<-
"main_reg" vr2_merged_acute1\$Region \<- "Hyaline Membrane" \\\

------------------------------------------------------------------------

# Transfer ROI to Molecule Data

To assess whether molecules are enriched in hyaline membrane zones.

\\{r} vr2_merged_acute1 \<- transferData(object = vr2_merged_acute1,
from = "Assay9", to = "Assay8", features = "Region") \\\

## Summary of Molecule Counts

\\{r} s2_summary_hyaline \<- Metadata(vr2_merged_acute1, assay =
"Xenium_mol") %\>% filter(gene %in% c("S2_N", "S2_orf1ab"), Region ==
"Hyaline Membrane") %\>% summarise(S2_N = sum(gene == "S2_N"), S2_orf1ab
= sum(gene == "S2_orf1ab"), ratio = sum(gene == "S2_N")/sum(gene ==
"S2_orf1ab")) %\>% as.matrix() s2_summary_hyaline \\\

🧠 **Interpretation**: `S2_N / S2_orf1ab` ratio informs us about **viral
replication**. A higher ratio indicates more **active replication**.

------------------------------------------------------------------------

# Interactive Visualization

Convert to AnnData/Zarr for `vitessce` or `rolv` visualization.

\\{r} as.AnnData(vr2_merged_acute1, assay = "Assay7", file =
"../data/ondisk/COVID19/vr2_merged_acute1_annotated_registered.zarr",
flip_coordinates = TRUE, create.ometiff = TRUE, name = "main", channel =
"H&E") \\\

------------------------------------------------------------------------

# Region-Based Comparison

Using interactive annotation, define **infected zones** and compare with
hyaline membranes.

\\{r} \# Manually annotate infected zones vr2_merged_acute1 \<-
annotateSpatialData(vr2_merged_acute1, assay = "Xenium_mol", label =
"Region2", use.image = TRUE, channel = "H&E", annotation_assay =
"InfectedAnnotation")

# Compare abundance

s2_summary_infected \<- Metadata(vr2_merged_acute1, assay =
"Xenium_mol") %\>% filter(gene %in% c("S2_N", "S2_orf1ab"), Region2 ==
"Infected") %\>% summarise(S2_N = sum(gene == "S2_N"), S2_orf1ab =
sum(gene == "S2_orf1ab"), ratio = sum(gene == "S2_N")/sum(gene ==
"S2_orf1ab")) %\>% as.matrix()

rbind(s2_summary_infected, s2_summary_hyaline) \\\

------------------------------------------------------------------------

# Hotspot Analysis

Identify **regions of high local molecule enrichment** using spatial
graph statistics.

\\{r} vr2_merged_acute1 \<- getSpatialNeighbors(vr2_merged_acute1, assay
= "Xenium_mol", group.by = "gene", group.ids = c("S2_N", "S2_orf1ab"),
method = "radius", radius = 50)

vr2_merged_acute1 \<- getHotSpotAnalysis(vr2_merged_acute1, assay =
"Xenium_mol", features = "gene", graph.type = "radius") \\\

------------------------------------------------------------------------

# Summary

This pipeline illustrates how **integrating multiple spatial data
types** can enhance biological insight:

-   **Image registration** aligns cells and morphology.
-   **ROI annotation** enables pathology-informed molecular analysis.
-   **Multi-layer visualization** reveals complex infection patterns.
-   **Ratio analysis** and **hotspot detection** highlight viral
    activity and tissue damage zones.

Together, they provide a powerful framework for spatial biology and
disease pathology studies.

------------------------------------------------------------------------

# Environment & Dataset Setup {#environment--dataset-setup}

## Run RStudio in Docker

Use a pre-built Docker image with VoltRon and all dependencies
pre-installed.

1.  Make sure [Docker
    Desktop](https://www.docker.com/products/docker-desktop) is
    installed.

2.  Pull the image:

``` bash
docker pull amanukyan1385/rstudio-voltron:main
```

3.  Start the container:

``` bash
docker run --rm -ti \
  -e PASSWORD=yourpassword \
  -p 8787:8787 \
  -v Path/to/your/dataset:/home/rstudio/project \
  amanukyan1385/rstudio-voltron:main
```

4.  Access RStudio:\
    <http://localhost:8787>
    -   Username: `rstudio`
    -   Password: `yourpassword` \# you can change your own password

You can also use Docker Desktop GUI if you prefer volume mounting.

------------------------------------------------------------------------

## Datasets

Download the complete dataset bundle used in this notebook:

[**Download all datasets (ZIP)**](#0) **- including datasets for other
markdown**

------------------------------------------------------------------------

### Original Dataset Sources

-   **Xenium – COVID-19 Lung**
    -   [Annotated
        RDS](https://bimsbstatic.mdc-berlin.de/landthaler/VoltRon/Multiomics/acutecase1_annotated.rds)
-   **Visium CytAssist – Breast Cancer**
    -   [Dataset](https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast)

------------------------------------------------------------------------

<nav style="background-color: #f8f9fa; padding: 10px; font-family: sans-serif; border-bottom: 1px solid #ddd;">

<a href="../index.html" style="margin-right: 20px; text-decoration: none;">
← Back to Note</a>
<a href="https://github.com/Dewey-Wang/Compgen-2025" target="_blank" style="text-decoration: none;">
View on GitHub</a>

</nav>
