---
title: "Spatial Transcriptomics - Xenium Analysis"
author: "Ding Yang Wang"
date: "2025-04-04"
output: html_document
output_dir: ../../../docs/module2
editor_options: 
  markdown: 
    wrap: 72
---

<nav style="background-color: #f8f9fa; padding: 10px; font-family: sans-serif; border-bottom: 1px solid #ddd;">

<a href="../index.html" style="margin-right: 20px; text-decoration: none;">
← Back to Note</a>
<a href="https://github.com/Dewey-Wang/Compgen-2025" target="_blank" style="text-decoration: none;">
View on GitHub</a>

</nav>

```{=html}
<style>
.center {
  display: flex;
  justify-content: center;
  align-items: center;
  text-align: center;
}
</style>
```

# Description

This R Markdown document performs a high-resolution **cell-level spatial
analysis** using **Xenium** breast cancer data and **VoltRon** package.
Unlike Visium (spot-based), Xenium captures **individual cells**,
allowing us to investigate **local microenvironments**, **immune-tumor
interactions**, and **expression hotspots**.

We perform a full analysis including: data filtering, disk-based
representation for large-scale handling, clustering, marker analysis,
and hotspot detection.

> If you want to reproduce the experiment: 👉 [Click here to the Docker
> and the data setup instructions](#environment--dataset-setup)

## Xenium Analysis – Cell-Level Spatial Transcriptomics

Compared to spot-based methods (e.g., Visium), Xenium captures
**individual cell** or even **molecule-level** information, enabling
high-resolution analysis of **cell identity**, **spatial interactions**,
and **functional zones** within tissues. This allows us to examine not
just spatial zones of gene expression but **individual cells and their
microenvironments**.

```{r setup, include=FALSE}
options(java.parameters = "-Xmx8g")
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

# Load Required Libraries

```{r load-libraries, message=FALSE}
library(HDF5Array)
library(ImageArray)
library(BPCells)
library(VoltRon)
library(rhdf5) 
library(Seurat)
library(patchwork)
library(ComplexHeatmap)
```

------------------------------------------------------------------------

# Data Import & Filtering

We start by importing the Xenium dataset and filtering out cells with
low expression counts (e.g., low RNA molecule counts per cell) to remove
noise. This helps reduce background signal and computational burden.

```{r import-xenium, eval = FALSE}
Xen_R1 <- importXenium("../workshop/data/BreastCancer/Xenium_R1/outs/", 
                       sample_name = "XeniumR1", 
                       overwrite_resolution = TRUE, 
                       resolution_level = 3)

# Filter cells with low molecule counts
Xen_R1 <- subset(Xen_R1, Count > 5)
```

------------------------------------------------------------------------

# On-Disk Representation

Xenium datasets are often large (especially in molecule-resolution), so
we store them on disk using an `HDF5VoltRon` format. Disk-backed formats
allow us to analyze large spatial datasets without crashing R due to
memory limits.

```{r, eval = FALSE}
# Save and reload Xenium object
Xen_R1_ondisk <- saveVoltRon(Xen_R1, 
                             format = "HDF5VoltRon", 
                             output = "../workshop/data/ondisk/Xen_R1", 
                             replace = TRUE)
Xen_R1_ondisk <- loadVoltRon("../workshop/data/ondisk/Xen_R1/")
```

```{r xenium-disk, echo=FALSE}
Xen_R1_ondisk <- readRDS("./RDS/Xen_R1_ondisk_1.rds")
```

------------------------------------------------------------------------

# Omic Profile Clustering

We normalize gene expression, perform dimensionality reduction (PCA +
UMAP), and cluster cells using SNN. The UMAP visualization shows
distinct clusters of cells based on expression, while the spatial plot
reveals **how those clusters organize within tissue**.

```{r, eval = FALSE}
Xen_R1_ondisk <- normalizeData(Xen_R1_ondisk, sizefactor = 1000)
Xen_R1_ondisk <- getPCA(Xen_R1_ondisk, dims = 20, overwrite = TRUE)
Xen_R1_ondisk <- getUMAP(Xen_R1_ondisk, dims = 1:20, overwrite = TRUE)
Xen_R1_ondisk <- getProfileNeighbors(Xen_R1_ondisk, dims = 1:20, method = "SNN")
Xen_R1_ondisk <- getClusters(Xen_R1_ondisk, resolution = 1.3, label = "Cluste rs", graph = "SNN")

# UMAP + spatial plots
p1 <- vrEmbeddingPlot(Xen_R1_ondisk, group.by = "Clusters", embedding = "umap", pt.size = 0.4, label = TRUE)
p2 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "Clusters", pt.size = 0.18)

(p1 | p2) 
```

```{r xenium-clustering, fig.align='center', fig.width=9, fig.height=4, echo=FALSE}
# UMAP + spatial plots
p1 <- vrEmbeddingPlot(Xen_R1_ondisk, group.by = "Clusters", embedding = "umap", pt.size = 0.4, label = TRUE)
p2 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "Clusters", pt.size = 0.18)

(p1 | p2) 
```

------------------------------------------------------------------------

# Marker Analysis & Cell Type Annotation

We identify marker genes and annotate each cluster based on known cell
types. This gives biological meaning to the clusters and helps define
what kinds of cells are where in the tissue.

```{r, eval = FALSE}
# Convert to Seurat for marker detection
Xen_R1$Clusters <- Xen_R1_ondisk$Clusters
Xen_R1_seu <- VoltRon::as.Seurat(Xen_R1, cell.assay = "Xenium", type = "image")
Idents(Xen_R1_seu) <- "Clusters"
Xen_R1_seu <- NormalizeData(Xen_R1_seu, scale.factor = 1000)
markers <- FindAllMarkers(Xen_R1_seu)

# Annotate clusters with known labels
annotations <- read.table("../workshop/data/BreastCancer/Xenium_R1/annotation.txt")[,1]
CellType <- factor(Xen_R1_ondisk$Clusters)
levels(CellType) <- annotations
Xen_R1_ondisk$CellType <- as.character(CellType)

# Visualize annotations
p1 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "CellType", pt.size = 0.18, alpha = 1)
p2 <- vrEmbeddingPlot(Xen_R1_ondisk, group.by = "CellType", embedding = "umap", pt.size = 0.4, label = TRUE)

# Save annotated object
Xen_R1_ondisk <- saveVoltRon(Xen_R1_ondisk)
(p1 | p2) 
```

```{r xenium-annotation, fig.align='center', fig.width=11, fig.height=4, echo=FALSE}
Xen_R1_ondisk <- readRDS("./RDS/Xen_R1_ondisk_2.rds")
# Visualize annotations
p1 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "CellType", pt.size = 0.18, alpha = 1)
p2 <- vrEmbeddingPlot(Xen_R1_ondisk, group.by = "CellType", embedding = "umap", pt.size = 0.4, label = TRUE)
(p1 | p2) 
```

------------------------------------------------------------------------

# Spatially Aware Clustering (Niche Clustering)

Now that we know cell types, we want to see how they're arranged in
tissue. Using **spatial neighborhoods**, we build a graph of cell-cell
proximity, compute the **abundance of cell types around each cell**,
normalize them (CLR), and perform **k-means clustering**.

These **niche clusters** reveal **tissue microenvironments** — like
immune zones or invasive tumor nests — based on **compositional
cell-type context**.

```{r, eval = FALSE}
# Build spatial neighbor graph
Xen_R1_ondisk <- getSpatialNeighbors(Xen_R1_ondisk, radius = 30, method = "radius")

# Build niche matrix
Xen_R1_ondisk <- getNicheAssay(Xen_R1_ondisk, label = "CellType", graph.type = "radius", overwrite =TRUE)

# Normalize + cluster niche profiles
vrMainFeatureType(Xen_R1_ondisk) <- "Niche"
Xen_R1_ondisk <- normalizeData(Xen_R1_ondisk, method = "CLR")
Xen_R1_ondisk <- getClusters(Xen_R1_ondisk, nclus = 9, method = "kmeans", label = "niche_clusters")

# Visualize
p1 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "niche_clusters", alpha = 1)
p2 <- vrHeatmapPlot(Xen_R1_ondisk, features = vrFeatures(Xen_R1_ondisk), group.by = "niche_clusters")
(p1 | p2) 
```

```{r xenium-niche, fig.align='center', fig.width=9, fig.height=3, echo=FALSE}
Xen_R1_ondisk <- readRDS("./RDS/Xen_R1_ondisk_3.rds")
p1 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "niche_clusters", alpha = 1)
p2 <- vrHeatmapPlot(Xen_R1_ondisk, features = vrFeatures(Xen_R1_ondisk), group.by = "niche_clusters")
library(gridExtra)
library(grid)
heatmap_grob <- grid.grabExpr(draw(p2))
grid.arrange(p1, heatmap_grob, ncol = 2)
```

> 💬 *Interpretation:* In the clustering plots, different **niche
> clusters** represent distinct **spatial microenvironments**. For
> example: cluster 2 (**ACTA2 and KRT-positive epithelial cells**)
> suggest structural zones, while immune-dense niches cluster
> separately. These patterns help us map **functional domains** within
> the tissue, guided by **cellular composition and spatial context**.

------------------------------------------------------------------------

# Subsetting Cell Types

To explore **immune infiltration** or **stromal interactions**, we
subset and visualize specific cell types. These plots help confirm
findings from the niche heatmaps.

```{r xenium-celltype,  fig.align='center', fig.width=9, fig.height=3}
p1 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "CellType", pt.size = 0.18, alpha = 1, 
              group.ids = c("ACTA2_myoepithelial", "KRT15_myoepithelial"))

p2 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "CellType", pt.size = 1, alpha = 1, 
              group.ids = c("CD4_TCells", "CD8_TCells", "BCells"), n.tile = 400)
(p1 | p2) 
```

> 🔍 Interpretation of Spatial & Niche Clustering Plots
>
> -   **ACTA2 / KRT15** myoepithelial cells form **structural
>     boundaries**, likely marking tumor edges or ductal structures.
> -   **CD4 / CD8 T cells and B cells** highlight areas of **immune
>     infiltration**, with CD8+ T cells suggesting cytotoxic responses
>     and B cells possibly forming lymphoid-like structures.

------------------------------------------------------------------------

# Hot Spot Analysis

This analysis detects **statistical hotspots** of a specific gene (e.g.
`PGR`), identifying areas where expression is **significantly elevated
in both a cell and its neighbors**.\
It uses **Getis-Ord statistics**, comparing local vs. global expression
means, and outputs per-cell p-values.

```{r, eval = FALSE}
# Build spatial neighbor graph for hotspot
Xen_R1_ondisk <- getSpatialNeighbors(Xen_R1_ondisk, method = "radius", radius = 15, graph.key = "radius_hot")

# Define RNA features as main
vrMainFeatureType(Xen_R1_ondisk) <- "RNA"

# Visualize gene
p1 <- vrSpatialFeaturePlot(Xen_R1_ondisk, features = "PGR", alpha = 1, background.color = "black", n.tile = 300)

# Run hotspot test
Xen_R1_ondisk <- getHotSpotAnalysis(Xen_R1_ondisk, features = "PGR", graph.type = "radius_hot", alpha.value = 0.001)

# Visualize significant regions
p2 <- vrSpatialFeaturePlot(Xen_R1_ondisk, features = "PGR_hotspot_stat", alpha = 1, background.color = "black", n.tile = 400)
p3 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "PGR_hotspot_flag", alpha = 1, background.color = "black", n.tile = 400)

p1 + p2 + p3
```

```{r xenium-hotspot,  fig.align='center', fig.width=12, fig.height=3, echo=FALSE}
Xen_R1_ondisk <- readRDS("./RDS/Xen_R1_ondisk_4.rds")

# Visualize gene
p1 <- vrSpatialFeaturePlot(Xen_R1_ondisk, features = "PGR", alpha = 1, background.color = "black", n.tile = 300)

# Visualize significant regions
p2 <- vrSpatialFeaturePlot(Xen_R1_ondisk, features = "PGR_hotspot_stat", alpha = 1, background.color = "black", n.tile = 400)
p3 <- vrSpatialPlot(Xen_R1_ondisk, group.by = "PGR_hotspot_flag", alpha = 1, background.color = "black", n.tile = 400)

p1 + p2 + p3
```

> 💬 *Explanation:* Hotspot analysis pinpoints **regions of biological
> activity**, such as hormone-active tumor zones or interferon
> responses. It complements clustering by analyzing **spatial patterns
> of single gene features**.

------------------------------------------------------------------------

# Summary

This Xenium pipeline provides a high-resolution, cell-level spatial
analysis of breast cancer tissue, including:

-   Cell identity classification
-   Microenvironment niche clustering
-   Detection of spatial gene expression hotspots

These steps help us understand **cellular organization, interactions,
and functional zones** — essential for cancer biology, immunology, and
tissue profiling.

------------------------------------------------------------------------

# Environment & Dataset Setup {#environment--dataset-setup}

## Run RStudio in Docker

Use a pre-built Docker image with VoltRon and all dependencies
pre-installed.

1.  Make sure [Docker
    Desktop](https://www.docker.com/products/docker-desktop) is
    installed.

2.  Pull the image:

``` bash
docker pull amanukyan1385/rstudio-voltron:main
```

3.  Start the container:

``` bash
docker run --rm -ti \
  -e PASSWORD=yourpassword \
  -p 8787:8787 \
  -v Path/to/your/dataset:/home/rstudio/project \
  amanukyan1385/rstudio-voltron:main
```

4.  Access RStudio:\
    <http://localhost:8787>
    -   Username: `rstudio`
    -   Password: `yourpassword` \# you can change your own password

You can also use Docker Desktop GUI if you prefer volume mounting.

------------------------------------------------------------------------

## Datasets

Download the complete dataset bundle used in this notebook:

[**Download all datasets (ZIP)**](#0) **- including datasets for other
markdown**

------------------------------------------------------------------------

### Original Dataset Sources

-   **Xenium – Breast Cancer**
    -   [Preview
        Dataset](https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast)

<nav style="background-color: #f8f9fa; padding: 10px; font-family: sans-serif; border-bottom: 1px solid #ddd;">

<a href="../index.html" style="margin-right: 20px; text-decoration: none;">
← Back to Note</a>
<a href="https://github.com/Dewey-Wang/Compgen-2025" target="_blank" style="text-decoration: none;">
View on GitHub</a>

</nav>
